---
title: |
  | DS311
  | Exploratory_Data_Analysis
subtitle: |
  | --------------------------------------------------
  | Introduction to EDA
  | --------------------------------------------------
output: pdf_document
# Author: Renzo Angelo A. Alasan
mainfont: Times New Roman
header-includes:
  - "\\usepackage{graphicx}"
  - "\\usepackage{titling}"
  - "\\pretitle{\\begin{center}\\LARGE\\includegraphics[width=4cm]{logo.png}\\\\[\\bigskipamount]}"
  - "\\posttitle{\\end{center}}"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 3: Lab Activity
Let’s put it all together with a hands-on **case study**. We’ll use the **Palmer Penguins** dataset (`penguins`
data frame from the `palmerpenguins` package). This real dataset contains measurements of three penguin
species (Adelie, Chinstrap, Gentoo) in Antarctica. The variables include species, island, sex, and numerical
measurements like flipper length, bill dimensions, and body mass.
**Scenario:** Imagine we are exploring this penguin data to understand their physical characteristics. We will
focus on the **body_mass_g** (body mass in grams) of the penguins and how it varies by species and overall.

```{r}
tinytex::install_tinytex()
```

